% ======================================================================
% STAGING DRAFT: MEASURED guardrail comparison (our matched protocol, frozen rows).
% All numbers here are measured by us on the identical 2,018 in-house + 2,020 novel frozen rows.
% Logprob guards -> threshold-free AUPRC + matched-FPR; binary-verdict guards -> native decision point.
% ======================================================================

\subsection{Specialized guards under our controlled protocol}
\label{sec:guardrail-measured}

The published numbers above are self-reported and not comparable across models (\S\ref{sec:guardrail-landscape}). To put a measured subset on a controlled footing, we re-evaluate the guards for which this repo has runnable local pipelines on the \emph{identical} frozen rows, using the same protocol as for our own guard. Llama Guard and ShieldGemma expose continuous unsafe scores and are therefore summarized by threshold-free AUPRC plus matched-FPR@0.10. Prompt Guard 2 is also continuous, but in this repo it is evaluated only on the injection/jailbreak subsets, which is its intended task. Qwen3Guard-Gen and WildGuard emit text verdicts, so we read them at their native decision point (\Cref{tab:gr-native}), exactly as we do for gpt-5.4-mini.

\begin{table*}[t]
\small\centering\setlength{\tabcolsep}{4pt}
\caption{Guards with a continuous score, \textbf{measured} on our rows: threshold-free AUPRC (in-house pooled / novel 3-set aggregate), matched-FPR@0.10 F1, batch=1 latency where measured, and BF16 VRAM. Our guard and its untuned base are repeated from \Cref{tab:auprc-poolings,tab:novel-auprc} for reference. Prompt Guard 2 is omitted here because it is evaluated only on the injection/jailbreak subsets and is discussed in the text.}
\label{tab:gr-auprc}
\begin{tabular}{lcccccc}
\toprule
System & Params & in-house & novel & mF1@.10 & lat.\ (ms) & VRAM \\
\midrule
Guard (SmolLM3-3B, tuned) & 3B & \textbf{0.844} & 0.781 & \textbf{0.581} & 124 & 6\,GB \\
Untuned base (SmolLM3-3B) & 3B & 0.696 & \textbf{0.886} & --- & 124 & 6\,GB \\
Llama-Guard-3-8B          & 8B & 0.750 & 0.833 & 0.475 & --- & 16\,GB \\
ShieldGemma-2B            & 2.6B & 0.712 & n/r & 0.464 & --- & 5\,GB \\
Llama-Guard-3-1B          & 1.1B & 0.639 & 0.701 & 0.360 & --- & 2.2\,GB \\
\bottomrule
\end{tabular}
\end{table*}

\begin{table*}[t]
\small\centering\setlength{\tabcolsep}{4pt}
\caption{Generative binary-verdict guards, \textbf{measured} at their native decision point on our rows (no tunable threshold, as for gpt-5.4-mini): F1 / FPR on the in-house pool and novel aggregate, plus FPR on the over-refusal-bait-heavy OR-Bench-hard set. Both over-block benign-adjacent prompts.}
\label{tab:gr-native}
\begin{tabular}{lccccc}
\toprule
System & Params & in-house F1/FPR & novel F1/FPR & OR-Bench-hard FPR & lat.\ (ms) \\
\midrule
Qwen3Guard-4B & 4B & 0.797 / 0.389 & 0.807 / 0.376 & 0.812 & 491 \\
WildGuard-7B  & 7B & 0.787 / 0.360 & 0.823 / 0.335 & 0.728 & 918 \\
gpt-5.4-mini  & --- & 0.784 / 0.321 & --- & --- & \textasciitilde{}512 \\
Guard (deployed pt) & 3B & 0.794 / 0.306 & --- & --- & 124 \\
\bottomrule
\end{tabular}
\end{table*}

Three findings matter. \textbf{(1) OOD ranking depends on what question we ask.} Among the general moderation guards with continuous scores, the novel-set AUPRC order is Llama-Guard-3-8B (0.833) $>$ our tuned guard (0.781) $>$ Llama-Guard-3-1B (0.701), while the untuned base remains strongest overall at 0.886. At the deployable matched-FPR@0.10 operating point, however, our 3B guard leads Llama-Guard-3-8B on F1 (0.581 vs.\ 0.475). This is the same ranking-fragility story as the main paper, now within the specialized-guard subset. \textbf{(2) Over-blocking is widespread among recall-first guards.} Qwen3Guard-4B and WildGuard-7B both post heavy false-positive rates on OR-Bench-hard (0.81 and 0.73), an independent cross-vendor confirmation of the over-refusal failure mode isolated in \S\ref{sec:mortgage-casestudy}. \textbf{(3) Prompt Guard 2 is a specialized detector, not a general moderator.} In this repo it is evaluated only on injection/jailbreak subsets, where it scores 0.903 AUPRC in-house and 0.625 on the novel subsets. It is therefore complementary to, not a replacement for, a general content guard.

\subsection{Recommendations}
\label{sec:recommendations}

Combining our controlled measurements with the published landscape and deployment profiles, and scoped to \emph{input-prompt} moderation (our evaluation setting; response- and tool-moderation are out of scope, \S\ref{sec:future}):

\begin{itemize}
  \item \textbf{Best overall guard (accuracy-first, resources permitting).} No single winner: Llama-Guard-3-8B is the strongest measured OOD ranker among off-the-shelf guards (novel AUPRC 0.833), and Qwen3Guard reports the strongest published generalist profile. The measured generative guards in this range are materially heavier than the 3B local guard: Qwen3Guard-4B and WildGuard-7B take about 0.5--0.9\,s per call and over-block, while Llama-Guard-3-8B requires 16\,GB BF16 memory. If a frontier API is acceptable, gpt-5.4-mini shows no statistically significant F1 difference from our guard.
  \item \textbf{Best lightweight base to fine-tune.} If in-house accuracy and a low false-positive operating point dominate, \textbf{Qwen3-4B} --- the highest measured in-house AUPRC (0.882) and matched-FPR@0.10 F1 (0.753) of the six bases, at 4B / 112\,ms (\Cref{tab:slm-sweep}); Qwen3-8B is statistically indistinguishable at double the memory. If efficiency dominates, any 1.5--3B base (SmolLM2, Qwen2.5, SmolLM3) reaches in-house AUPRC 0.82--0.85 at a fraction of the VRAM. Because tuning \emph{converges} out-of-distribution behavior across bases (\Cref{sec:cross-family}), the base barely affects OOD robustness --- so choose on in-house accuracy, cost, and license, not on hoped-for generalization; and where OOD robustness is paramount, keep the untuned base in a base$\oplus$tuned ensemble (\Cref{sec:ensemble}).
  \item \textbf{Best prompt-injection / jailbreak detector.} Prompt Guard 2-86M --- 0.903 measured in-house-injection AUPRC, 0.998 published jailbreak AUC, 0.17\,GB BF16, sub-100\,ms; pair it with a content guard rather than using it alone.
  \item \textbf{Best content-category moderation.} ShieldGemma (published OpenAI-Mod F1 0.82) for policy-category outputs; but on our hardened domain set every small guard over-blocks (\S\ref{sec:mortgage-casestudy}), so category moderation still needs an operating-point audit.
  \item \textbf{Best performance/efficiency balance.} Our laptop-trained 3B guard: among the locally measured general content guards it leads in-house (AUPRC 0.844) and at matched-FPR (F1 0.581) at 6\,GB / 124\,ms, and shows no statistically significant F1 difference from gpt-5.4-mini; the untuned base or a base$\oplus$tuned ensemble (\S\ref{sec:ensemble}) is preferable when out-of-distribution robustness dominates.
  \item \textbf{Best for real-time production.} For pure injection/jailbreak gating, Prompt Guard 2 (encoder, single forward pass, sub-100\,ms). For general prompt moderation, our 3B guard is the fastest locally measured general-content option in this repo at 124\,ms batch=1; Llama-Guard-3-1B is smaller in VRAM (2.2\,GB) but its batch=1 latency was not measured here. The measured 4--7B generative guards (0.5--0.9\,s) are hard to justify inline. Qwen3Guard-Stream (published) is the one design here explicitly built for streaming.
\end{itemize}
