\section{Limitations, the evidence ledger, and the validation roadmap}
\label{sec:honesty}

This section is the report's honest accounting. Everything above is a set of \emph{measurements}; this section states precisely what those measurements do and do not license you to conclude, records the provenance and label tier of each body of evidence in one unified ledger, and lays out the concrete steps that would upgrade each estimate into something stronger. Two rules are enforced without exception and should be read as the spine of the whole report: \textbf{(1)} retrospective, estimation-only evidence is \emph{never} pooled with external or prospective evidence, and \textbf{(2)} no number anywhere is promoted into a causal, universal, or fair-lending claim.

\begin{background}{The difference between ``what we measured'' and ``what you may conclude''}
Most of this report reports \emph{paired differences} on a \emph{fixed} set of four models: for each checkpoint we compare the guard to its own base on identical rows with identical scoring. That design is unusually clean for what it targets --- it removes the model-to-model confound that ordinary leaderboards suffer --- but it buys that cleanliness by giving up breadth. A paired difference on four hand-chosen checkpoints tells you what happened \emph{to these four models on these rows}; it does not, by itself, tell you what will happen to a fifth model, on a production traffic mix, or why. The bootstrap intervals quantify sampling noise \emph{conditional on this panel and these benchmarks}; they are not confidence statements about a population of models or prompts, and they are not hypothesis tests. Keeping that distinction visible is the entire point of this section.
\end{background}

\subsection{What these results do NOT establish}
\label{sec:not-establish}

\paragraph{Cross-cutting claims we do not make.} The following limits apply to \emph{every} act, regardless of which axis is in view.

\begin{enumerate}[leftmargin=1.6em,itemsep=2pt]
  \item \textbf{No causal claim.} We observe that fine-tuning \emph{co-occurs} with specialization (Act~I, \Cref{sec:actI}) and that composition \emph{co-occurs} with transfer recovery (Act~II, \Cref{sec:actIII}). We do not isolate a mechanism. In particular, the reading that ``stronger bases specialize more'' --- suggested by the ordering from SmolLM2-1.7B ($+0.0400$ transfer change) down to Qwen3-4B ($-0.1499$) --- is a \emph{hypothesis drawn from four points}, not a demonstrated law. It is also mechanically entangled with the metric: because the paired change is $\Delta=\mathrm{AP}_{\text{SFT}}-\mathrm{AP}_{\text{base}}$ and every SFT guard converges to $\approx0.98$ represented-source AP regardless of where its base started ($0.4524{\to}0.9806$ for SmolLM2 up to $0.8855{\to}0.9837$ for Qwen3-4B), $\Delta$ is arithmetically coupled to the base level. A negative correlation between base AP and $\Delta$ is partly a definitional artifact and cannot be read as evidence of a behavioral tendency.

  \item \textbf{No population or universal claim.} The panel is a \emph{fixed, purposively chosen} set of four checkpoints drawn from only \emph{two model lineages} (Qwen: Qwen2.5-1.5B, Qwen3-4B; and the SmolLM family: SmolLM2-1.7B, SmolLM3-3B). Lineage, scale, and native alignment recipe are therefore confounded: any ``size effect'' we might read off four points is inseparable from ``which of two families it came from.'' Nothing here estimates a distribution over models, and the hierarchical-bootstrap intervals (e.g.\ the aggregate transfer change $\TransferDelta$~[$\TransferCILower,\TransferCIUpper$]) are explicitly \emph{conditional on this panel}, not population intervals.

  \item \textbf{Heterogeneous native policies are not controlled.} Each base checkpoint arrives with its \emph{own} native safety training and implicit policy. A paired base$\to$SFT delta therefore mixes ``what our fine-tune did'' with ``what this vendor's alignment already did,'' and the four bases do not share a common definition of \code{unsafe}. This is a feature for the paired design (we always compare a model to itself) but a limit on interpretation: the cross-checkpoint spread partly reflects incompatible starting policies, not just differing tunability.

  \item \textbf{Not confirmatory --- the benchmarks were inspected during development.} The 1{,}200-row training manifest and the transfer suite were visible to the researcher while the method was being built. ``Held out'' throughout this report means \emph{dataset-held-out} (rows/sources not used in training), \emph{not} sealed-from-the-researcher. This makes Acts~I and II \emph{retrospective and estimation-only}. No pre-registration protects them; the honest status is ``a reproducible characterization of this panel,'' not ``a confirmed finding.'' The one exception is the starting-type adaptation study (\Cref{sec:adaptation}), whose estimands, decision rules, and non-inferiority margin were locked in a committed claim registry (\code{artifacts/starting\_type\_adaptation\_v1/protocol/claim\_registry.json}, hashed in the release lock) \emph{before} any score existed --- it is confirmatory (though still conditional on the fixed model panel).

  \item \textbf{Decontamination against the v2 transfer suite is asserted, not yet audited.} The manifest was decontaminated during construction and is reported as clean-v2, but the \emph{formal} overlap audit (n-gram / near-duplicate check of the training manifest against the current v2 transfer benchmarks) is still pending. If residual train/transfer overlap exists, it would \emph{inflate} measured transfer --- i.e.\ it would make specialization look milder than it is --- so the pending audit is a caveat in the direction of caution.

  \item \textbf{Single recipe, single manifest.} Act~I uses exactly one LoRA configuration (rank~32, $\alpha{=}64$, dropout~0.05 on \code{q,k,v,o,gate,up,down}; 300 steps; lr~$2\!\times\!10^{-4}$ cosine, warmup~0.03; effective batch~4; \code{max\_len}~1024; seeds 42--46 with data order fixed at seed~42) on one frozen manifest. The specialization signature is a property of \emph{this recipe on this data}; a different rank, step budget, or data mixture could move the represented/transfer trade-off, and we do not sweep them.

  \item \textbf{Balanced-prevalence evaluation overstates production precision.} The transfer regime is scored on a balanced pool (\TransferPositiveN{} unsafe vs.\ \TransferNegativeN{} safe rows) and the represented regime is near-balanced (\RepPositiveN{} vs.\ \RepNegativeN{}). Real inbound traffic is overwhelmingly benign, so unsafe prompts are \emph{rare}. Average precision and any precision-at-threshold measured on a balanced set are optimistic relative to a deployment where the negative class dominates: at low base rates the same ranking yields far lower precision. This is now made quantitative rather than left as a caveat --- \Cref{sec:actI-prevalence} and \Cref{fig:prevalence} recompute $\mathrm{AP}(\pi_+)$ exactly from the ranking (\Cref{eq:ap-prevalence}): the base transfer guards fall from $0.79$--$0.94$ balanced to $\approx0.10$--$0.54$ at $1\%$ prevalence (\Cref{fig:prevalence}), and the guard ordering itself re-spaces as positives get rare. Read every AP here as a \emph{ranking} quality on a balanced pool, not as a deployed precision.

  \item \textbf{Ranking recovery is not threshold transfer.} All AP-based results are threshold-free; deployment is not. At a calibration-selected operating point the gap is stark: SFT lifts represented-source recall from $\RepBaseTPRPct$\% to $\RepSFTTPRPct$\% but raises the transfer false-alarm rate from $\TransferBaseFPRPct$\% to $\TransferSFTFPRPct$\% (macro) --- and pooled FPR moves further, $\TransferBasePooledFPRPct$\%$\to\TransferSFTPooledFPRPct$\% --- while HarmBench recall \emph{falls} from $\HarmBenchBaseRecallPct$\% to $\HarmBenchSFTRecallPct$\% (\Cref{tab:sensitivity}). Composition recovers rank but its realized transfer FPR at a 5\% target is $\PilotCompositionTransferFPRPct$\% --- better than SFT's $\PilotSFTTransferFPRPct$\% yet still above target (\Cref{tab:pilot-operating}). A better AP does not hand you a transferable cutoff. The size of these blow-ups is telling: a Gaussian-tail calculation maps the macro FPR shift ($8.1\%\!\to\!15.5\%$) to only $\approx0.4$ standard deviations of calibration drift, and the pooled shift ($4.3\%\!\to\!17.0\%$) to $\approx0.8$~SD --- sub-sigma drifts that no realistic calibration protocol excludes, because a fixed cutoff's false-alarm rate is \emph{exponentially} sensitive to drift at the operating quantile.

  \item \textbf{Prompt-only scope.} We classify the \emph{incoming request} (binary \code{safe}/\code{unsafe}) via the single-token $z_{\text{unsafe}}-z_{\text{safe}}$ head. We do not evaluate response classification, multi-turn context, tool-call arguments, or streamed content. Guards that read the assistant's \emph{output} or a full dialogue are a different task and are out of scope; nothing here should be read as a claim about them.
\end{enumerate}

\paragraph{Act~I (SFT specialization) --- what it does not establish.} The headline aggregate transfer change of $\TransferDelta$ is \emph{not} a clean ``fine-tuning barely hurts transfer.'' It pools opposing per-checkpoint effects (Qwen2.5-1.5B $-0.0389$~[$-0.0829,+0.0062$]; SmolLM2-1.7B $+0.0400$~[$+0.0003,+0.0776$]; SmolLM3-3B $-0.0869$~[$-0.1114,-0.0613$]; Qwen3-4B $-0.1499$~[$-0.1963,-0.1050$]); the panel average is small precisely because it averages a gain against three losses. The specialization plane (\Cref{fig:plane}) shows $\SpecializationSeedCount$ of $\TotalSeedCount$ (checkpoint, seed) points in the specialize quadrant and $\UniformGainSeedCount$ in uniform gain --- a \emph{tendency on this panel}, not a universal direction, and one seed-family (SmolLM2) runs the other way.

\paragraph{Act~II (composition) --- what it does not establish.} Composition (\Cref{eq:comp}) recovers transfer relative to SFT ($\PilotTransferDeltaSFT$~[$\PilotTransferDeltaSFTLow,\PilotTransferDeltaSFTHigh$]) at a small represented cost ($\PilotRepDeltaSFT$~[$\PilotRepDeltaSFTLow,\PilotRepDeltaSFTHigh$]), landing above the base on transfer ($\PilotCompositionTransfer$ vs.\ $\PilotBaseTransfer$). But: \textbf{(a)} it is \emph{not} Pareto dominance --- versus base it is heterogeneous, helping the weaker bases (SmolLM2 $+0.067$, Qwen2.5 $+0.036$ vs.\ base transfer) while \emph{hurting} the strongest, Qwen3-4B ($-0.030$); \textbf{(b)} the mechanism is unproven (the ensemble/diversity account is a motivation, not a proof); \textbf{(c)} one control is \emph{run} and one is still missing --- we have run the equal-inference-cost SFT+SFT baseline (\Cref{tab:sftsft}: base+SFT beats SFT+SFT on all four checkpoints, attributing the recovery to ``keeping the base'' rather than to ``ensembling anything''), but not a true weight-space WiSE-FT rescoring \citep{wortsman2022wiseft}, so we cannot say which family recovers more transfer here; and \textbf{(d)} the logit-average and convex-weight variants (transfer $\PilotLogitTransfer$ for logit-average) were \emph{visible during development} and are reported only as non-promotable ablations. Whether composition also recovers transfer for guards tuned with \emph{other} objectives is left open.

\paragraph{Act~III, mortgage --- a measuring stick, not a legal finding.} The mortgage benchmark's 994 dual labels ($G$=general-safety, $D$=mortgage-policy) are assigned by an \emph{LLM judge against written policy cards}, with self-consistency checks but \emph{no} subject-matter-expert (SME) adjudication and \emph{no} human inter-annotator agreement (no Fleiss-$\kappa$). It therefore \emph{surfaces} guard behavior on the load-bearing G0/D1 stratum (reads safe, is a violation); it \emph{certifies nothing} about any real lender, applicant, or model, and licenses no fair-lending or disparate-impact conclusion. Four further limits bound it: \textbf{(i)} the \textbf{G1/D0 quadrant is empty} (0 rows), so ``looks unsafe yet is policy-compliant'' --- the over-refusal-on-compliant-requests failure mode --- is \emph{unmeasurable} here; \textbf{(ii)} several strata are \emph{small} (G1/D1 = 42 rows; private\_test = 95 rows; 39 protected minimal pairs / 78 rows), so per-stratum estimates are noisy; \textbf{(iii)} the zero-shot bases rank policy violations only \emph{moderately} (AP$\cdot$D $0.67$--$0.85$; \Cref{tab:baseline}), and because asking for discrimination or fraud already reads as unsafe to a general model, $G$ and $D$ are only \emph{partially} orthogonal --- much of the subtle G0/D1 stratum stays unresolved by ranking; \textbf{(iv)} the fixed-threshold operating point is deliberately \emph{not} tabulated because it is knife-edge for these clustered-score guards (its G0/D1 catch count swung by more than 50 rows across library versions) --- itself a finding about unreliable threshold transfer, not a number to deploy. The protected-pair gap $\Delta_{\mathrm{context}}$ is a fairness \emph{signal} (Qwen3-4B invariant at $0.000$; Qwen2.5-1.5B $0.183$; max observed $0.51$), not a fairness \emph{verdict}.

\paragraph{Act~III, ExpGuard --- external, single-label, base-only, never pooled.} The finance/health/law replication uses \emph{external, expert-annotated} labels (\code{prompt\_label}) over 2{,}275 rows --- a strictly stronger labeling tier than the mortgage LLM-judge. It tests whether the specialization/transfer pattern \emph{recurs} across regulated verticals, but it is \textbf{single-label} (not the mortgage dual $G\times D$ construct), so it is a breadth/transfer probe, not a compliance construct. The four-checkpoint \emph{base} result is complete (\Cref{tab:expguard}) and reproducible from committed per-row scores; the tuned (base-vs-SFT) comparison on ExpGuard is future work. Because ExpGuard's labels come from a different source and tier, its numbers are \emph{never} averaged, ranked, or otherwise pooled with the mortgage LLM-judge numbers or with the retrospective panel.

\begin{takeaway}{The one-line version}
We measured paired base-to-tuned changes on a fixed two-lineage panel, on benchmarks we had seen, at balanced prevalence, using ranking metrics --- plus one depth domain with LLM-judge labels and one external breadth domain with expert labels. That supports ``on this panel, tuning specializes and composition recovers some transfer.'' It does \emph{not} support any causal, universal, deployment, or fair-lending claim.
\end{takeaway}

\subsection{The unified evidence ledger}
\label{sec:ledger}

The report carries evidence at \emph{different tiers}, and the honesty of the synthesis depends on never letting a weaker tier borrow credibility from a stronger one. Two dimensions matter. The first is the \textbf{evidence flavor}: is the number \emph{retrospective / estimation-only} (measured after the fact on inspected data, conditional on a fixed panel) or does it come from an \emph{external, independently annotated} source? The second is the \textbf{label tier}, in decreasing strength: \emph{SME-adjudicated with reported inter-annotator agreement} (the gold standard --- we have \emph{none} of this yet); \emph{external expert-annotated} (ExpGuard); and \emph{LLM-judge, policy-card-consistent} (the mortgage benchmark --- self-consistent against written cards, but no human). \Cref{tab:ledger} records, for each body of evidence, its flavor, its label tier, what it establishes, and what would upgrade it. The governing rule is stated once and applied everywhere: \textbf{the three columns of flavor are never pooled}. Retrospective panel numbers (Acts~I--II), the LLM-judge mortgage numbers (Act~III depth), and the external expert-annotated ExpGuard numbers (Act~III breadth) are reported side by side but never averaged into a single headline.

\begin{table}[htbp]\centering\footnotesize
\caption{The unified evidence ledger. Each row records one body of evidence, its flavor and label tier, what it establishes, and the concrete upgrade that would strengthen it. The three flavors are never pooled; no row is promoted to a causal, universal, or fair-lending claim.}
\label{tab:ledger}
\setlength{\tabcolsep}{4pt}\renewcommand{\arraystretch}{1.15}
\begin{tabular}{@{}p{0.15\textwidth}p{0.20\textwidth}p{0.28\textwidth}p{0.29\textwidth}@{}}
\toprule
Body of evidence & Flavor \& label tier & Establishes (conditional on panel/data) & Principal limits $\to$ upgrade \\
\midrule
Act~I: SFT specialization (\Cref{tab:primary,tab:sensitivity}, \Cref{fig:plane}) &
Retrospective, estimation-only; dataset-held-out; benchmarks inspected in dev &
On this panel, SFT lifts represented-source AP ($\RepDelta$) but not transfer ($\TransferDelta$); $\SpecializationSeedCount/\TotalSeedCount$ seeds specialize &
Two lineages; single recipe; balanced prevalence; $\Delta$ coupled to base; v2 decontamination audit pending $\to$ prospective uninspected cohort + overlap audit \\
Act~II: composition (\Cref{tab:pilot-summary,tab:pilot-models,tab:pilot-operating}, \Cref{eq:comp}) &
Retrospective, estimation-only; same panel &
Vs.\ SFT, recovers transfer ($\PilotTransferDeltaSFT$) at small represented cost ($\PilotRepDeltaSFT$); above base on transfer &
Not Pareto (hurts Qwen3-4B); ablations dev-visible; SFT+SFT control run (\Cref{tab:sftsft}, base+SFT wins on all four); WiSE-FT rescoring still open $\to$ add weight-space control \\
Act~III depth: mortgage $G\times D$ (\Cref{tab:composition,tab:baseline}, \Cref{fig:pipeline}) &
LLM-judge, policy-card-consistent; \emph{not} SME; 994 rows &
A dual-labeled, HMDA-grounded stick for the G0/D1 stratum + a protected-pair fairness signal ($\Delta_{\mathrm{context}}$) &
No human agreement; empty G1/D0; small strata; knife-edge threshold $\to$ SME adjudication + Fleiss-$\kappa$; populate G1/D0 \\
Act~III breadth: ExpGuard (\Cref{tab:expguard}) &
External, expert-annotated; single-label; base-only (complete); 2{,}275 rows &
Whether the specialization/transfer pattern recurs across finance/health/law with expert labels &
Single-label (not dual $G\times D$); base-only $\to$ add the tuned (base-vs-SFT) comparison; dual-label expert construct \\
\bottomrule
\end{tabular}
\end{table}

\begin{background}{Why ``never pool'' is not just bookkeeping}
If you average an LLM-judge score (which can inherit the judge model's blind spots) with an expert-annotated score (which cannot), the blended number is neither. Worse, it launders the weaker label's uncertainty into the stronger one and produces a headline no single method would support. Keeping the flavors in separate columns means a reader can always ask ``how was this labeled?'' and get a straight answer --- and it means the mortgage benchmark's honest status (a measuring stick, not a certification) is never quietly upgraded by proximity to the expert-annotated set.
\end{background}

\subsection{The validation roadmap}
\label{sec:roadmap}

Each limitation above has a matching upgrade. The roadmap is ordered roughly by how much it would strengthen the central claims, and every item names the specific weakness it closes.

\begin{enumerate}[leftmargin=1.9em,itemsep=3pt]
  \item \textbf{Lock a genuinely uninspected prospective cohort} (closes ``not confirmatory,'' Acts~I--II). Pre-register the estimands and decision rules, then evaluate on data neither the researcher nor the manifest builder has seen. This is the single upgrade that would move Acts~I and II from ``retrospective characterization of this panel'' to a confirmatory finding; the adaptation study (\Cref{sec:adaptation}) already applies this discipline to the starting-type question, and the remaining step is to extend it to Acts~I--II's core specialization estimands on genuinely uninspected data.
  \item \textbf{Complete the v2 decontamination audit} (closes the pending-audit caveat). Run the formal n-gram / near-duplicate overlap check of the 1{,}200-row manifest against the v2 transfer suite and report the residual overlap; any leakage found would revise the transfer estimates \emph{downward} (i.e.\ specialization would look sharper).
  \item \textbf{Add the remaining composition control} (closes Act~II's mechanism gap). The equal-inference-cost \textbf{SFT+SFT} baseline is now run (\Cref{tab:sftsft}: base+SFT beats two-adapter ensembling on every checkpoint, so the recovery is the base's doing); the open item is a true weight-space \textbf{WiSE-FT} rescoring \citep{wortsman2022wiseft} to compare output-space against weight-space interpolation on identical rows.
  \item \textbf{SME-adjudicate a stratified mortgage subset and report Fleiss-$\kappa$} (closes the mortgage label-tier gap). Have qualified subject-matter experts independently re-label a stratified subset (weighted toward the G0/D1 and protected-pair rows), report inter-annotator agreement, and reconcile against the LLM-judge labels; this is what would let any mortgage number graduate from ``measuring stick'' toward an audit-grade instrument.
  \item \textbf{Populate the empty G1/D0 quadrant} (closes an unmeasurable failure mode). Construct grounded ``looks unsafe yet policy-compliant'' requests so over-refusal on legitimate mortgage traffic becomes measurable rather than absent by construction.
  \item \textbf{Evaluate at production prevalence with a cost-weighted operating point} (closes ``balanced prevalence overstates precision''). Re-score at realistic (heavily benign) base rates and report precision and a cost-weighted threshold, so the deployed-precision story is not read off a balanced pool.
  \item \textbf{Broaden the panel beyond two lineages} (closes the lineage/scale confound). Add checkpoints from additional model families and a wider size range so that ``stronger bases specialize more'' can be tested rather than inferred from four coupled points.
  \item \textbf{Extend ExpGuard to the tuned comparison and build dual-labeled finance/health constructs with expert sign-off} (closes the breadth gap). The four-checkpoint \emph{base} ExpGuard eval is complete (\Cref{tab:expguard}); add the tuned (base-vs-SFT) comparison, and extend the mortgage-style dual-label $G\times D$ construct into finance and health with expert adjudication --- one domain built in depth plus verticals built for breadth.
  \item \textbf{Score gated / commercial guards and extend beyond prompt-only} (closes the non-commercial-data and scope limits). Once licenses are accepted, evaluate gated open guards (e.g.\ Llama~Guard~3, WildGuard) on the same rows and scorer; and extend the task beyond input-prompt classification to response and multi-turn moderation. Note that several training sources are non-commercial / gated, which constrains redistribution --- raw third-party rows are referenced by pinned identifier + revision + content hash, never redistributed.
\end{enumerate}

\begin{takeaway}{The disciplined next step}
The honest upgrade path is not a more confident headline; it is a \emph{prospectively locked, decontaminated, appropriately controlled} re-run on genuinely uninspected data, with SME-adjudicated labels where compliance is at stake. Until then, every claim in this report stands exactly as written: a reproducible, estimation-only characterization of one fixed panel, plus one LLM-judge depth domain and one external expert-annotated breadth domain --- reported honestly, and never pooled.
\end{takeaway}
